The evolution of gene expression in the term placenta of viviparous mammals

### 2015-10-13 manuscript/manuscript.Rnw
include supplemental material external references
add go analysis of core placenta transcriptome

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
\usepackage[plain]{fancyref}
\renewcommand*{\fref}[1]{\Fref{#1}}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
\pagestyle{fancy}

@@

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")

@@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset")))>>=
for (a in paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset"))) {
    load(a)
}
setkey(genes.to.tree,"gene_id")
star.logs[,value:=gsub("%$","",value)]
star.logs.wide <- 
    data.table(dcast(star.logs,species+file~field))
setnames(star.logs.wide,
         "Average input read length",
         "input.read.length")
setnames(star.logs.wide,
         "Number of input reads",
         "input.reads")
setnames(star.logs.wide,
         "Uniquely mapped reads number",
         "unique.map.reads")
@

@@

\begin{figure}
<<core.placenta.transcriptome,fig.width=8,fig.height=12,echo=FALSE>>=
combined.fpkm.wide <- 
    data.table(dcast(combined.fpkm,
                     human_name~species,
                     fun.aggregate=sum,
                     value.var="mean_fpkm"))
temp <- 
    combined.fpkm.wide[,median_expression := apply(combined.fpkm.wide[,-1,with=FALSE],1,median)]
temp <- 
    combined.fpkm.wide[,min_expression := apply(combined.fpkm.wide[,-1,with=FALSE],1,min)]
setkey(combined.fpkm.wide,"human_name")
core.placenta.transcriptome.long <- 
    data.table(melt(combined.fpkm.wide[min_expression >= 10,
                                       ][order(-median_expression)
                                         ][!(human_name %in% 
                                                 housekeeping.genes.superset[,gene_short_name])],
                    id.vars="human_name",
                    variable.name="Species",
                    value.name="FPKM"
                    ))
temp <- core.placenta.transcriptome.long[,human_name:=
                                     reorder(human_name,
                                             combined.fpkm.wide[core.placenta.transcriptome.long[,human_name],median_expression])]
rm(temp)

print(ggplot(core.placenta.transcriptome.long[!Species %in% c("median_expression","min_expression"),],
             aes(y=human_name,x=capfirst(Species)))
      + geom_tile(aes(fill=log10(FPKM)))
      + scale_x_discrete(expand=c(0,0))
      + scale_y_discrete(expand=c(0,0))
      + scale_fill_gradient(low="white",
                            high="blue",
                            limits=c(0,4),
                            guide=guide_colorbar(title=expression(log[10](FPKM)))
                            )
      + ylab("Human Ortholog Name")
      + xlab("Species")
      + theme_grey()
      + theme(legend.position="top",
              axis.ticks = element_blank(), 
              axis.text.x = element_text(angle = 340, hjust = 0.3, vjust=0.8,face="italic"))
      )

# 
# print(xtable(combined.fpkm.wide[min_expression > 10,
#                                 ][order(-median_expression)
#                                   ][!(human_name %in% 
#                                           housekeeping.genes.superset[,gene_short_name])][1:75,],
#              digits=0
#              ),
#       include.rownames=FALSE,
#       booktabs=TRUE,
#       floating=FALSE
#       )
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  genes expressed (\ac{FPKM}$\ge 10$) in the placenta of all organisms
  studied with housekeeping
  genes\cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
  removed.}
\label{fig:core_placenta_transcriptome}
\end{figure}

>>>>

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{wrapfig}
% \usepackage{txfonts}
\usepackage[plain]{fancyref}
\renewcommand*{\fref}[1]{\Fref{#1}}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*{supplemental_material}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\renewcommand\cite[1]{\autocite{#1}}
\pagestyle{fancy}

@@

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -1, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")

@@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","palcenta_go_results_mf")))>>=
for (a in paste0("../data/",c("combined_fpkm","gene_trees",
                              "housekeeping_genes_superset",
                              "placenta_core_transcriptome",
                              "placenta_go_results_bp",
                              "placenta_go_results_mf"
                              ))) {
    load(a)
}
setkey(genes.to.tree,"gene_id")
star.logs[,value:=gsub("%$","",value)]
star.logs.wide <- 
    data.table(dcast(star.logs,species+file~field))
setnames(star.logs.wide,
         "Average input read length",
         "input.read.length")
setnames(star.logs.wide,
         "Number of input reads",
         "input.reads")
setnames(star.logs.wide,
         "Uniquely mapped reads number",
         "unique.map.reads")
@

@@

\begin{figure}
<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
print(ggplot(core.placenta.transcriptome.long[!Species %in% c("median_expression","min_expression"),],
             aes(y=human_name,x=capfirst(Species)))
      + geom_tile(aes(fill=log10(FPKM)))
      + scale_x_discrete(expand=c(0,0))
      + scale_y_discrete(expand=c(0,0))
      + scale_fill_gradient(low="white",
                            high="blue",
                            limits=c(0,4),
                            guide=guide_colorbar(title=expression(log[10](FPKM)))
                            )
      + ylab("Human Ortholog Name")
      + xlab("Species")
      + theme_grey()
      + theme(legend.position="top",
              axis.ticks = element_blank(), 
              axis.text.x = element_text(angle = 340, hjust = 0.3, vjust=0.8,face="italic"))
      )
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
  genes expressed (\ac{FPKM}$ \ge 10$) in the placenta of all organisms
  studied with housekeeping
  genes \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
  removed.}
\label{fig:core_placenta_transcriptome}
\end{figure}

<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=
go.results <- rbindlist(list(placenta_go_results_mf[1:10,],
                             placenta_go_results_bp[1:10,]))
go.results <- go.results[,Annotated:=NULL]
setnames(go.results,"classic","p value")
setnames(go.results,"classic.fdr","FDR")
setnames(go.results,"GO.ID","GO ID")
setnames(go.results,"Term","GO Term")
print(xtable(go.results,
             digits=2,
             display=c("d","s","s","d","f","e","e","s"),
             caption="Top go MF results",
             label="tab:core_placenta_go_terms"),
      include.rownames=FALSE,
      floating=TRUE,
      booktabs=TRUE,
      size="\\tiny"
      )
@

The
\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\fref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the transcriptome if an ortholog of that gene was expressed with
$\mathrm{FPKM} \ge 10$ in all species studied and was not a
housekeeping gene
\cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
\textit{ANXA1} is involved in the modulation of the inflammatory
response \DLA{It does more stuff; read the huthcinson2011 review and
  fill this in}
%
Another annexin, \textit{ANXA2} 
%
\textit{TIMP3}, which may be down-regulated in pre-eclampsia
\cite{Ma.ea2014:Down-regulationofTIMP3leads} (though another study
\cite{Xiang.ea2013:PromoterhypomethylationofTIMP3} claims
up-regulation)
